Evaluating a lightweight transcriptome assembly pipeline on two closely related ascidian species

### 2014-09-16 manuscript/main.tex
final pass minor fixes

\documentclass[fleqn,10pt]{wlpeerj}
\title{Assembly on Cloud 9: Assembly metric assessment using Transcriptomes of closely related Ascidian species}

@@

\keywords{Assembly, Cloud computing, Ascidians, Next-gen sequencing}

@@

\begin{table}[ht]
\centering
\includegraphics[scale=1.0]{Table1.pdf}
\caption{\textbf{Digital normalized reads.} The number of reads sequenced before and after digital normalization are shown for each lane of sequencing. The percentage of total reads kept after digital normalization is shown in bold. \textit{M. occulta} had approximately \mytilde237 million reads and was reduced to 91 million reads, a 60\% reduction. \textit{M. oculata} had 150 million reads and reduced by 77\% to \mytilde50 million reads. }
\label{fig:Table1.pdf}
\end{table}

@@

The difference in assembly time and memory between diginorm and raw reads was not as large when using the Trinity assembler. Diginorm completed its assemblies several hours faster than assembling raw reads, \mytilde15 hours compare to \mytilde26 hours for \textit{M. oculata} and \mytilde24 hours compared to \mytilde39 hours for \textit{M. occulta}. \textit{M. occulta} normalized reads assembled with 23.17 GB of RAM versus 34.14 GB of RAM for the unnormalized reads (Figure~\ref{fig:time_and_memory.pdf}).

\begin{figure}[tbp]
\centering
\includegraphics[scale=0.5]{figure_1_time_and_memory.pdf}
\caption{\textbf{Wall time and memory requirements for assemblies.} Wall time (left) in hours to complete the diginorm (DN) and raw read (RAW) assemblies for both species and assemblers. Oases assembled multiple k's, 21 $\leq$ k $\leq$ 35 opposed to Trinity that assembles a single k. This is one reason the assembly times differed. (right) Shows the memory used to assemble each of the transcriptomes. M. oculata (ocu) transcriptomes assemble in less time than M. occulta (occ) because they have fewer lanes of reads to assemble. In all cases diginorm required less time and memory to complete the assembly.}
\label{fig:time_and_memory.pdf}
\end{figure}

\subsection{Assembly statistics varied by preprocessing approach and assembler}
Oases run with the diginormed reads yielded fewer total transcripts than Oases run with the unnormalized reads. The \textit{M. oculata} diginorm assembly produced 300 fewer transcripts, and the \textit{M. occulta} diginorm assembly produced 227 less transcripts (Table~\ref{fig:Table2.pdf}). Digital normalization had the opposite affect when using Trinity for assembly, increasing the total number of assembled transcripts by 7,840 for \textit{M. oculata} and 9,197 for \textit{M. occulta}.

@@

\subsection{Read mapping shows high inclusion of reads in the assembled transcriptomes}
	We mapped the quality-filtered reads to the assembled transcriptomes to evaluate their inclusiveness. The F+3 stage of reads from \textit{M. occulta} had the lowest percentage of mapped reads, with the Oases unnormalized assembly mapping only 49\% of the reads, and the Trinity unnormalized assembly mapping 67\% (Figure 3\ref{fig:figure_3_mocc_mapping}). This was an isolated case: all other Oases assemblies contained at least 75\% of the reads for each time point and the Trinity assemblies contained at least 93\% of the reads for each time point. Trinity raw read assemblies tended to contain slightly more reads than the diginorm assemblies, while the opposite was true for Oases; however, in no case did the raw-reads assembly differ from the diginorm assemblies in more than 3\% of their read content.
\setcounter{subfigure}{0}
\begin{figure}[tbp]
	\begin{center}
	\subfigure[]{%
                \label{fig:figure_3_mocc_mapping}
                \includegraphics[width=0.4\textwidth]{figure_3_mocc_mapping.pdf}
	}%
	\subfigure[]{%
                \label{fig:figure_3_mocu_mapping}
                \includegraphics[width=0.4\textwidth]{figure_3_mocu_mapping.pdf}
	}%
	\end{center}
	\caption{\textbf{Read mapping.} Unnormalized reads were mapped back to each of the assemblies to determine the inclusion of reads in the assembly. ~\ref{fig:figure_3_mocc_mapping}\textit{M. occulta} first round of gastrulation reads (f+3), showed the lowest mapping quality for all assemblies, with the lowest being raw Oases at 48.57\%. \textit{M. occulta} f+3 is the only case were mapping is less than 74\% and the only case where DN Trinity mapped more reads than Raw Trinity. ~\ref{fig:figure_3_mocu_mapping}\textit{M. oculata} unnormalized Oases performed the worst, with Trinity assemble having the best mappings. Trinity assemblies have more mapped reads than Oases for all conditions, having at least 93\% read mapping for both species. Raw Trinity typically mapped slightly more reads than DN, and the opposite occurs for Oases, with DN having more reads mapped to its assembly.}
	\label{fig:read_mapping}
\end{figure}

@@

Of these 178 nucleotide sequences, 8 of them are \textit{M. occulta} sequences and 15 of them are \textit{M. oculata} sequences. All \textit{M. occulta} assemblies recovered all 8 of the NCBI \textit{M. occulta} sequences with a 94\% or greater accuracy. \textit{M. oculata} assemblies recovered \textit{M. oculata} transcripts at a 93\% accuracy as well. \textit{M. occulta} assemblies produced the lowest completeness of the two species, 41\% and 43\% for unnormalized Oases and Diginorm Oases respectively, and 75\% for both Trinity assemblies. \textit{M. oculata} assemblies produced more complete transcripts 66, 75, 86, and 83 percent for unnormalized Oases, Diginorm Oases, unnormalized Trinity and Diginorm Trinity respectively.

\subsection{Both unnormalized and normalized assemblies recovered many of the same transcripts}
We next evaluated the two diginorm and unnormalized assemblies against one another to test whether either method missed significant portions of the transcriptome assembled by the other. We used BLAT to compare unnormalized and diginorm assemblies in both directions. In \textit{M. occulta}, both methods recovered at least 93\% of the transcripts, with Trinity diginorm recovering \mytilde99\% of Trinity's unnormalized assembly. \textit{M. oculata} assemblies showed high overlap as well, all recovering greater than 98\% of each other with the exception of diginorm Oases recovering 94\% of unnormalized Oases assembly.

\subsection{Homology search against the \textit{Ciona} proteome shows similar recovery of ascidian genes across assemblies}
We next used \textit{Ciona intestinalis} to evaluate the completeness of our transcriptomes. \textit{C. intestinalis} has an assembled genome that is well annotated and is the closest available genome to the Molgulids. \textit{C. intestinalis} has a genome of ~160 Mb and contains \mytilde16,000 genes \citep{satoh_surfing_2005}. A total of 13, 835 (86\%) of the \textit{C. intestinalis} proteins found in NCBI had hits in the \textit{M. occulta} transcriptomes (Figure 5), with 2,288 genes (14\%) having no hits due presumably to either lack of expression, high divergence, or loss \textit{M. occulta}. When comparing transcripts excluded by either diginorm or unnormalized reads for all assemblies, the unnormalized read assemblies produced an additional 0.04\% hits to \textit{C. intestinalis} and there was additional 0.03\% for the diginorm assemblies. There was little difference between the assemblies when compared to \textit{C. intestinalis}, with 99\% of the \textit{C. intestinalis} genes being found in all \textit{M. occulta} assemblies (Figure 4a). Eighty-six percent of the \textit{C. intestinalis} proteins had matches in the \textit{M. occulta} and \textit{M. oculata} assemblies with less than 1\% difference in presence between the several assemblies (Figure 4b).

 \setcounter{subfigure}{0}
\begin{figure}[tbp]
	\begin{center}
	\subfigure[\textit{ M. occulta}]{%
                \label{fig:figure_5_mocc_blast_overlap}
                \includegraphics[width=0.4\textwidth]{figure_5_mocc_blast_overlap.pdf}
	}%
	\subfigure[\textit{ M. oculata}]{%
                \label{fig:figure_5_mocu_blast_overlap}
                \includegraphics[width=0.4\textwidth]{figure_5_mocu_blast_overlap.pdf}
	}%
	\end{center}
	\caption{\textbf{Gene recovery, raw reads versus Normalized.} Gene homologue with C. intestinalis via BLAST for M. occulta (left) and M. oculata (right). Each oval represent the total number of homologs sequences recovered. In both species the Trinity assembler assembled more homologous sequences. There was almost complete overlap in homology for both assemblers and both assembly conditions.}
	\label{fig:overlap}
\end{figure}

>>>>

\documentclass[fleqn,10pt]{wlpeerj}
\title{Evaluating a lightweight transcriptome assembly pipeline on two closely related ascidian species}

@@

\keywords{Assembly, Cloud computing, Ascidians, Next-generation sequencing}

@@

\begin{table}[ht]
\centering
\includegraphics[scale=1.0]{Table1.pdf}
\caption{\textbf{Digitally normalized reads.} The number of reads sequenced before and after digital normalization are shown for each lane of sequencing. The percentage of total reads kept after digital normalization is shown in bold. \textit{M. occulta} had approximately \mytilde237 million reads and was reduced to 91 million reads, a 60\% reduction. \textit{M. oculata} had 150 million reads and reduced by 77\% to \mytilde50 million reads. }
\label{fig:Table1.pdf}
\end{table}

@@

The difference in assembly time and memory between diginorm and raw reads was not as large when using the Trinity assembler. Diginorm completed its assemblies several hours faster than assembling raw reads, \mytilde15 hours compared to \mytilde26 hours for \textit{M. oculata} and \mytilde24 hours compared to \mytilde39 hours for \textit{M. occulta}. \textit{M. occulta} normalized reads assembled with 23.17 GB of RAM versus 34.14 GB of RAM for the unnormalized reads (Figure~\ref{fig:time_and_memory.pdf}).

\begin{figure}[tbp]
\centering
\includegraphics[scale=0.5]{figure_1_time_and_memory.pdf}
\caption{\textbf{Wall time and memory requirements for assemblies.} Wall time (left) in hours to complete the diginorm (DN) and raw read (RAW) assemblies for both species and assemblers. Oases assembled multiple k's, 21 $\leq$ k $\leq$ 35 opposed to Trinity that uses only a single k. This is one reason the assembly times differed. (right) Shows the memory used to assemble each of the transcriptomes. M. oculata (ocu) transcriptomes assemble in less time than M. occulta (occ) because they have fewer lanes of reads to assemble. In all cases diginorm required less time and memory to complete the assembly.}
\label{fig:time_and_memory.pdf}
\end{figure}

\subsection{Assembly statistics varied by preprocessing approach and assembler}
Oases run with the diginormed reads yielded fewer total transcripts than Oases run with the unnormalized reads. The \textit{M. oculata} diginorm assembly produced 300 fewer transcripts, and the \textit{M. occulta} diginorm assembly produced 227 fewer transcripts (Table~\ref{fig:Table2.pdf}). Digital normalization had the opposite affect when using Trinity for assembly, increasing the total number of assembled transcripts by 7,840 for \textit{M. oculata} and 9,197 for \textit{M. occulta}.

@@

\subsection{Read mapping shows high inclusion of reads in the assembled transcriptomes}
	We mapped the quality-filtered reads to the assembled transcriptomes to evaluate their inclusiveness. The F+3 stage of reads from \textit{M. occulta} had the lowest percentage of mapped reads, with the Oases unnormalized assembly mapping only 49\% of the reads, and the Trinity unnormalized assembly mapping 67\% (Figure 3\ref{fig:figure_3_mocc_mapping}). This was an isolated case: all other Oases assemblies contained at least 75\% of the reads for each time point and the Trinity assemblies contained at least 93\% of the reads for each time point. Trinity raw read assemblies tended to contain slightly more reads than the diginorm assemblies, while the opposite was true for Oases; however, in no case did the raw-reads assembly differ from the diginorm assemblies in more than 3\% of their read content.
\setcounter{subfigure}{0}
\begin{figure}[tbp]
	\begin{center}
	\subfigure[]{%
                \label{fig:figure_3_mocc_mapping}
                \includegraphics[width=0.4\textwidth]{figure_3_mocc_mapping.pdf}
	}%
	\subfigure[]{%
                \label{fig:figure_3_mocu_mapping}
                \includegraphics[width=0.4\textwidth]{figure_3_mocu_mapping.pdf}
	}%
	\end{center}
	\caption{\textbf{Read mapping.} Unnormalized reads were mapped back to each of the assemblies to determine the inclusion of reads in the assembly. ~\ref{fig:figure_3_mocc_mapping}\textit{M. occulta} first round of gastrulation reads (f+3), showed the lowest mapping quality for all assemblies, with the lowest being raw Oases at 48.57\%. \textit{M. occulta} f+3 is the only case were mapping is less than 74\% and the only case where DN Trinity mapped more reads than Raw Trinity. ~\ref{fig:figure_3_mocu_mapping}\textit{M. oculata} unnormalized Oases performed the worst, with Trinity assemble having the best mappings. Trinity assemblies have more mapped reads than Oases for all conditions, with at least 93\% read mapping for both species. Raw Trinity typically mapped slightly more reads than DN, and the opposite occurs for Oases, with DN having more reads mapped to its assembly.  Note that the Y axis starts at 45\%.}
	\label{fig:read_mapping}
\end{figure}

@@

Of these 178 nucleotide sequences, 8 of them are \textit{M. occulta} sequences and 15 of them are \textit{M. oculata} sequences. All \textit{M. occulta} assemblies recovered all 8 of the NCBI \textit{M. occulta} sequences with a 94\% or greater accuracy. \textit{M. oculata} assemblies recovered \textit{M. oculata} transcripts at a 93\% accuracy as well. \textit{M. occulta} assemblies produced the lowest completeness of the two species, 41\% and 43\% for unnormalized Oases and diginorm Oases respectively, and 75\% for both Trinity assemblies. \textit{M. oculata} assemblies produced more complete transcripts 66, 75, 86, and 83 percent for unnormalized Oases, Diginorm Oases, unnormalized Trinity and Diginorm Trinity respectively.

\subsection{Both unnormalized and normalized assemblies recovered many of the same transcripts}
We evaluated the two diginorm and unnormalized assemblies against one another to test whether either method missed significant portions of the transcriptome assembled by the other. We used BLAT to compare unnormalized and diginorm assemblies in both directions. In \textit{M. occulta}, both methods recovered at least 93\% of the transcripts, with Trinity diginorm recovering \mytilde99\% of Trinity's unnormalized assembly. \textit{M. oculata} assemblies showed high overlap as well, all recovering greater than 98\% of each other with the exception of diginorm Oases recovering 94\% of unnormalized Oases assembly.

\subsection{Homology search against the \textit{Ciona} proteome shows similar recovery of ascidian genes across assemblies}
We used \textit{Ciona intestinalis} to evaluate the completeness of our transcriptomes. \textit{C. intestinalis} has an assembled genome that is well annotated and is the closest available genome to the Molgulids. \textit{C. intestinalis} has a genome of ~160 Mb and contains \mytilde16,000 genes \citep{satoh_surfing_2005}. A total of 13, 835 (86\%) of the \textit{C. intestinalis} proteins found in NCBI had hits in the \textit{M. occulta} transcriptomes (Figure 5), with 2,288 genes (14\%) having no hits due presumably to either lack of expression, high divergence, or loss \textit{M. occulta}. When comparing transcripts excluded by either diginorm or unnormalized reads for all assemblies, the unnormalized read assemblies produced an additional 0.04\% hits to \textit{C. intestinalis} and there was additional 0.03\% for the diginorm assemblies. There was little difference between the assemblies when compared to \textit{C. intestinalis}, with 99\% of the \textit{C. intestinalis} genes being found in all \textit{M. occulta} assemblies (Figure 4a). Eighty-six percent of the \textit{C. intestinalis} proteins had matches in the \textit{M. occulta} and \textit{M. oculata} assemblies with less than 1\% difference in presence between the several assemblies (Figure 4b).

 \setcounter{subfigure}{0}
\begin{figure}[tbp]
	\begin{center}
	\subfigure[\textit{ M. occulta}]{%
                \label{fig:figure_5_mocc_blast_overlap}
                \includegraphics[width=0.4\textwidth]{figure_5_mocc_blast_overlap.pdf}
	}%
	\subfigure[\textit{ M. oculata}]{%
                \label{fig:figure_5_mocu_blast_overlap}
                \includegraphics[width=0.4\textwidth]{figure_5_mocu_blast_overlap.pdf}
	}%
	\end{center}
	\caption{\textbf{Gene recovery, raw reads versus normalized.} Gene homologue with C. intestinalis via BLAST for M. occulta (left) and M. oculata (right). Each oval represent the total number of homologs sequences recovered. In both species the Trinity assembler assembled more homologous sequences. There was almost complete overlap in homology for both assemblers and both assembly conditions.}
	\label{fig:overlap}
\end{figure}
